##  Killing aphids


 An experiment was designed to examine how well the insecticide rotenone kills
aphids that feed on the chrysanthemum plant called *Macrosiphoniella sanborni*. 
The explanatory variable is the log concentration (in milligrams per litre) of the
insecticide. At each of the five different concentrations,
approximately 50 insects were exposed. The number of insects exposed
at each concentration, and the number killed, are shown below.


```

Log-Concentration   Insects exposed    Number killed   
0.96                       50              6               
1.33                       48              16              
1.63                       46              24              
2.04                       49              42              
2.32                       50              44              

```




(a) Get these data into R. You can do this by copying the data
into a file and reading that into R (creating a data frame), or you
can enter the data manually into R using `c`, since there are
not many values. In the latter case, you can create a data frame or
not, as you wish. Demonstrate that you have the right data in R.
 
Solution

 
There are a couple of ways. My
current favourite is the `tidyverse`-approved
`tribble` method. A `tribble` is a 
"transposed `tibble`", in which you copy and paste the data,
inserting column headings and commas in the right places. The
columns don't have to line up, since it's the commas that
determine where one value ends and the next one begins:

```{r aphids-1 }
dead_bugs <- tribble(
  ~log_conc, ~exposed, ~killed,
  0.96, 50, 6,
  1.33, 48, 16,
  1.63, 46, 24,
  2.04, 49, 42,
  2.32, 50, 44
)
dead_bugs
```

    

Note that the last data value has no comma after it, but instead has
the closing bracket of `tribble`.  

You can have extra spaces if you wish. They will just be ignored.
If you are clever in R
Studio, you can insert a column of commas all at once (using
"multiple cursors").
I used to do it like this. I make vectors of each column using `c` and then glue the columns together into a data frame:

```{r aphids-2 }
log_conc <- c(0.96, 1.33, 1.63, 2.04, 2.32)
exposed <- c(50, 48, 46, 49, 50)
killed <- c(6, 16, 24, 42, 44)
dead_bugs2 <- tibble(log_conc, exposed, killed)
dead_bugs2
```

     

The values are correct --- I checked them.

Now you see why `tribble` stands for "transposed tibble": if you want to construct a data frame by hand, you have to work with columns and then glue them together, but `tribble` allows you to work "row-wise" with the data as you would lay it out on the page.

The other obvious way to read the data values without typing them is to copy
them into a file and read *that*. The values as laid out are
aligned in columns. They might be separated by tabs, but they are
not. (It's hard to tell without investigating, though a tab is by
default eight spaces and these values look farther apart than that.)
I copied them into a file `exposed.txt` in my current folder
(or use `file.choose`):

```{r aphids-3 }
bugs2 <- read_table("exposed.txt")
bugs2
```

 

This didn't quite work: the last column `Number killed` got
split into two, with the actual number killed landing up in
`Number` and the column `killed` being empty. If you
look at the data file, the data values for `Number killed` are
actually aligned with the word `Number`, which is why it came
out this way. Also, you'll note, the column names have those
"backticks" around them, because they contain illegal characters
like a minus sign and spaces. Perhaps a good way to
pre-empt^[My daughter learned the word pre-empt because we  like to play a bridge app on my phone; in the game of bridge, you  make a pre-emptive bid when you have no great strength but a lot of  cards of one suit, say seven, and it won't be too bad if that suit  is trumps, no matter what your partner has. If you have a weakish hand with a lot of cards in one suit, your opponents are probably going  to be able to bid and make something, so you pre-emptively bid first  to try and make it difficult for them.]  all these problems is to
make a copy of the data file with the illegal characters replaced by
underscores, which is my file `exposed2.txt`:

```{r aphids-4 }
bugs2 <- read_table("exposed2.txt")
bugs2
```

 
You might get a fictitious fourth column, but if you do, you can safely ignore it the rest of the way. We have all our actual data.
We'd have to be careful with Capital Letters
this way,  but it's definitely good.

You may have thought that this was a lot of fuss to make about reading
in data, but the point is that data can come your way in lots of
different forms, and you need to be able to handle whatever you
receive so that you can do some analysis with it.
 
$\blacksquare$ 

(b) <a name="part:expo">*</a> Looking at the data, would you expect there to be a
significant effect of log-concentration? Explain briefly.
 
Solution


The numbers of insects killed goes up *sharply* as the
concentration increases, while the numbers of insects exposed
don't change much. So I would expect to see a strong, positive
effect of concentration, and I would expect it to be strongly
significant, especially since we have almost 250 insects altogether.
 
$\blacksquare$


(c) Run a suitable logistic regression, and obtain a summary of
the results.

 
Solution


This is one of those where each row of data is more than one individual (there were about 50 insects at each concentration), so the response variable needs to have two columns: one of how many insects were killed at that dose, and the other of how many insects survived. This is done using `cbind`. We don't actually have a number of insects that survived, but we know how many were exposed, so we can work out how many survived by doing a subtraction inside the `cbind`:

```{r aphids-7 }
bugs.1 <- glm(cbind(killed, exposed - killed) ~ log_conc, family = "binomial", data = dead_bugs)
summary(bugs.1)
```

   
Extra: In the original part (c), I had you make the response matrix first, which is how I used to teach it, but I came to realize that this way plays better with `marginaleffects` because all the things in the `glm` line are in the dataframe.
 
$\blacksquare$

(d) Does your analysis support your answer to (<a href="#part:expo">here</a>)?
Explain briefly.

 
Solution


That's a *very* small P-value, $1.1\times 10^{-15}$, on
`log_conc`, so there is no doubt that concentration has an
effect on an insect's chances of being killed. This is exactly what
I guessed in (<a href="#part:expo">here</a>), which I did before looking at the
results, honest!

You might also note that the Estimate for `log_conc` is positive and, by the standards of these things, pretty large in size, so the probability of being killed goes up pretty sharply as log-concentration increases.
 
$\blacksquare$

(e) Obtain predicted probabilities of an insect's being killed at
each of the log-concentrations in the data set. 

 
Solution

Use a `newdata` that is the original dataframe, and display only the columns of interest:

```{r}
cbind(predictions(bugs.1, newdata = dead_bugs)) %>% 
  select(log_conc, exposed, killed, estimate)
```


 
The advantage of this is that you can
compare the observed with the predicted. For example, 44 out of 50
insects were killed at log-dose 2.32, which is a proportion of 0.88,
pretty close to the prediction of 0.91.

(f) Obtain a plot of the predicted probabilities of an insect's being killed, as it depends on the log-concentration.

This is `plot_predictions`, pretty much the obvious way:

```{r}
plot_predictions(model = bugs.1, condition = "log_conc")
```

The thing I always forget is that the name of the explanatory variable has to be *in quotes*, which is not what we are used to from the `tidyverse`. The odd thing on the $y$-axis is our response variable; the predictions are the probability of being killed, which we could make our $y$-axis actually say, like this:

```{r}
plot_predictions(model = bugs.1, condition = "log_conc") +
  labs(y = "Prob(killed)")

```


As the log-concentration goes up, you can see how clearly the probability of the aphid being killed goes up. The confidence band is narrow because there is lots of data (almost 250 aphids altogether), and the confidence interval is much higher at the end than at the beginning, which is why the P-value came out so tiny.

$\blacksquare$ 

(g) People in this kind of work are often interested in the
"median lethal dose". In this case, that would be the
log-concentration of the insecticide that kills half the
insects. Based on your predictions, roughly what do you think the
median lethal dose is?

 
Solution

This is probably easier to see from the graph, though the $x$-scale doesn't make it very easy to be accurate: look across to where the prediction crosses the probability of 0.5, which is somewhere between a log-concentration of 1.5 and 1.6.

If you go back and look at the predictions, the log-concentration of 1.63 is predicted to kill just over half
the insects, so the median lethal dose should be a bit less than
1.63. It should not be as small as 1.33, though, since that
log-concentration only kills less than a third of the insects. So I
would guess somewhere a bit bigger than 1.5. Any guess somewhere in
this ballpark is fine: you really cannot be very accurate.


Extra: this is kind of a strange prediction problem, because we know what
the *response* variable should be, and we want to know what the
explanatory variable's value is. Normally we do predictions the
other way around.^[This kind of thing is sometimes called an inverse prediction.]
So the only way to get a more accurate figure is
to try some different log-concentrations, and see which one gets
closest to a probability 0.5 of killing the insect.

Something like this would work:

```{r}
new <- datagrid(model = bugs.1, log_conc = seq(1.5, 1.63, 0.01))
cbind(predictions(bugs.1, newdata = new))
```


The closest one of these to a probability of 0.5 is 0.4971, which goes
with a log-concentration of 1.57: indeed, a bit bigger than 1.5 and a
bit less than 1.63, and close to what I read off from my graph. The `seq` in the construction of the new
data frame is "fill sequence": go from 1.5 to 1.63 in steps of
0.01. We are predicting for values of `log_conc` that we chose, so the way to go is to make a new dataframe with `datagrid` and then feed that into `predictions` with `newdata`.

Now, of course this is only our "best guess", like a single-number
prediction in regression. There is uncertainty attached to it (because
the actual logistic regression might be different from the one we
estimated), so we ought to provide a confidence interval for it. But
I'm riding the bus as I type this, so I can't look it up right now.

Later: there's a function called `dose.p` 
in `MASS` that appears to do this:

```{r aphids-11 }
lethal <- dose.p(bugs.1)
lethal
```

 

We have a sensible point estimate (the same 1.57 that we got by hand),
and we have a standard error, so we can make a confidence interval by
going up and down twice it (or 1.96 times it) from the estimate. The
structure of the result is a bit arcane, though:

```{r aphids-12 }
str(lethal)
```

 

It is what R calls a "vector with attributes". To get at the pieces and calculate the interval, we have to do something like this:

```{r aphids-13 }
(lethal_est <- as.numeric(lethal))
(lethal_SE <- as.vector(attr(lethal, "SE")))
```

 

and then make the interval:

```{r aphids-14 }
lethal_est + c(-2, 2) * lethal_SE
```

 

1.47 to 1.68.

I got this idea from page 4.14 of
[link](http://www.chrisbilder.com/stat875old/schedule_new/chapter4.doc). I
think I got a little further than he did. An idea that works more
generally is to get several intervals all at once, say for the
"quartile lethal doses" as well:

```{r aphids-15 }
lethal <- dose.p(bugs.1, p = c(0.25, 0.5, 0.75))
lethal
```

 

This looks like a data frame or matrix, but is actually a 
"named vector", so `enframe` will get at least some of this and turn
it into a genuine data frame:

```{r aphids-16 }
enframe(lethal)
```

 

That doesn't get the SEs, so we'll make a new column by grabbing the "attribute" as above:

```{r aphids-17 }
enframe(lethal) %>% mutate(SE = attr(lethal, "SE"))
```

 

and now we make the intervals by making new columns containing the lower and upper limits:

```{r aphids-18 }
enframe(lethal) %>%
  mutate(SE = attr(lethal, "SE")) %>%
  mutate(LCL = value - 2 * SE, UCL = value + 2 * SE)
```

 

Now we have intervals for the median lethal dose, as well as for the doses that kill a quarter and three quarters of the aphids.


$\blacksquare$
 



