## The Cross-City Line

When I went to university (in Birmingham, England, a long time ago), I was very excited because I would be travelling to campus by train. My journey was on the Cross-City Line, a metro-type service with lots of stops short distances apart, but run in those days by diesel trains (the electrification came later). 

```{r cross-city-1, echo=FALSE}
my_url <- "http://ritsokiguess.site/datafiles/cross-city.csv"
```


A list of the stations on the line is in `r my_url`. There is one column in the data file, called `station`. We are going to draw a map of these.

(a) Read in and display (some of) the station names.

Solution

Nothing terribly unexpected here:

```{r cross-city-2}
stations <- read_csv(my_url)
stations
```

$\blacksquare$

(b) In preparation for geocoding, create a second column in the dataframe that consists of the station names with "station UK" on the end. (This is to improve the chances of the geocoder finding the actual railway station.)

Solution

I wrote this back into the original dataframe. Create a new one if you prefer:

```{r cross-city-3}
stations %>% 
  mutate(longname = str_c(station, " station UK")) -> stations
stations
```

$\blacksquare$

(c) Look up the longitudes and latitudes of all the stations, organizing your dataframe so that they are visible.

Solution

Two steps: the first is to do the geocoding, and the second is to disentangle what comes back.

First, then:

```{r cross-city-4}
stations %>% 
  rowwise() %>% 
  mutate(ll = list(geocode_OSM(longname))) -> stations
stations
```

The longitudes and latitudes are hidden in the list-column that I called `ll`, so the second step is to get them out:

```{r cross-city-5}
stations %>% unnest_wider(ll) %>% 
  unnest_wider(coords) -> stations
stations
```

The two `unnest_wider`s are because the longitudes and latitudes are hidden inside a thing called `coords` which is itself nested within `ll`. Do the first `unnest_wider`, and see what you have. This will tell you that you need to do another one. 

The values seem reasonable; this is the UK, most of which is slightly west of the Greenwich meridian, and the latitudes look sensible given that the UK is north of southern Ontario.

$\blacksquare$

(d) Make a Leaflet map of the stations. Use circle markers or the "pin" markers as you prefer.


Solution

I used the pin markers (with `addMarkers`), though `addCircleMarkers` is as good. The code for drawing the map is always the same; the work here is in setting up the geocoding:

```{r cross-city-6}
leaflet(data = stations) %>% 
  addTiles() %>% 
  addMarkers(lng = ~x, lat = ~y)
```

You might find that some of the markers are off a straight line through Birmingham. In these cases, the geocoder found a different place with (apparently) the same name, or a name that was close enough to similar. This seems to be different each time you run it, so you can try again.

This (mostly) seems to extend across the city of Birmingham, as it should. There are quite a lot of stations, so the pins overlap each other. Zoom in to see them in a bit more detail, or zoom out to orient yourself if your UK geography needs some work.

$\blacksquare$

(e) Zoom in to see whether the geocoding did indeed find each of the stations. Comment briefly on what you find.

Solution

The stations go south to north, so the most southerly one you find should be Redditch and the most northerly is Lichfield Trent Valley. 

If you zoom in enough, you'll see where the railway line goes (a grey line). The points seem to be mainly close to it. But if you zoom in a bit more, some of the pins are right on the railway, and some of them are a bit off, because the geocoder found the centre of the place (or something else named after the place) rather than its railway station. For example, when I did it, Gravelly Hill station was right where it should be, but Aston was not.^[If you're a soccer fan, this Aston is what Aston Villa is named after. See if you can find the team's stadium Villa Park on your map, which is actually closer to Witton station on another line.]

Extra: `geocode_OSM` uses a free geocoder called Nominatim. This has some options. The defaults are to return only the first "hit", and to return just the coordinates and the "bounding box". These can be changed. Let's see what we can find for Aston:

```{r cross-city-7}
tibble(where = "Aston UK") %>% 
  mutate(info = list(geocode_OSM(where, return.first.only = FALSE,
                            details = TRUE))) -> d
d         
```
There are now 10 things returned. Let's unnest this and see what we have:

```{r cross-city-8}
d %>% unnest(info) %>% 
  unnest_wider(info)
```

There are 10 locations it found matching "Aston UK", and for each of those there is the information you see, a total of 12 columns' worth in addition to the name of the place we are looking up. Perhaps the most interesting are the columns `class` and `type` near the end (keep scrolling right):

```{r cross-city-9}
d %>% unnest(info) %>% 
  unnest_wider(info) %>% 
  select(where, class, type)

```

You can see which one is the railway station. 

This makes me think that with sufficient patience we could do this for all our places, and pick out the one that is the station:

```{r cross-city-10}
stations <- read_csv(my_url)
stations %>% 
  mutate(longname = str_c(station, " UK")) %>% 
  rowwise() %>% 
  mutate(ll = list(geocode_OSM(longname, 
                   return.first.only = FALSE,
                   details = TRUE))) -> stations2
```

```{r}
stations2
```

The number in the `ll` column tells you how many things the geocoder found that match the input `longname`. One of each of these is, we hope, a railway station.

```{r cross-city-11}
stations2 %>% unnest(ll) %>% 
  unnest_wider(ll, names_sep = "_") %>% 
  select(station, ll_coords, ll_class, ll_type) %>% 
  filter(ll_class == "railway", ll_type == "station") %>% 
  unnest_wider(ll_coords) -> d
d
```


If you want to see how this works, run it one line at a time.

We almost got there. We're missing University^[The station is called just University, not Birmingham University, which makes it hard to find.] and Lichfield City stations, but it looks as if we got the rest:

```{r cross-city-12}
leaflet(data = d) %>% 
  addTiles() %>% 
  addMarkers(lng = ~x, lat = ~y)
```

If you zoom in, you'll see that the ones we got are all the actual stations, and not the area from which the station takes its name.

$\blacksquare$

