##  The effects of Substance A


 In a dose-response experiment, animals (or
cell cultures or human subjects) are exposed to some toxic substance,
and we observe how many of them show some sort of response. In this
experiment, a mysterious Substance A is exposed at various doses to
100 cells at each dose, and the number of cells at each dose that
suffer damage is recorded. The doses were 10, 20, ... 70 (mg), and
the number of damaged cells out of 100 were respectively 10, 28, 53,
77, 91, 98, 99.



(a) Find a way to get these data into R, and show that you have
managed to do so successfully.


Solution


There's not much data here, so we don't need to create a file,
although you can do so if you like (in the obvious way: type the
doses and damaged cell numbers into a `.txt` file or
spreadsheet and read in with the appropriate `read_`
function). 
Or, use a `tribble`:

```{r doseresponse-1 }
dr <- tribble(
  ~dose, ~damaged,
  10, 10,
  20, 28,
  30, 53,
  40, 77,
  50, 91,
  60, 98,
  70, 99
)
dr
```

       
Or, make a data frame with the values typed in:

```{r doseresponse-2 }
dr2 <- tibble(
  dose = seq(10, 70, 10),
  damaged = c(10, 28, 53, 77, 91, 98, 99)
)
dr2
```

 

`seq` fills a sequence "10 to 70 in steps of 10", or you can
just list the doses.


For these, find a way you like, and stick with it.

    

$\blacksquare$

(b) Would you expect to see a significant effect of dose for
these data?
Explain briefly.


Solution


The number of damaged cells goes up sharply as the dose goes up
(from a very small number to almost all of them). So I'd expect
to see a strongly significant effect of dose. This is far from
something that would happen by chance.
 

$\blacksquare$

(c) Fit a logistic regression modelling the probability of a
cell being damaged as it depends on dose. Display the
results. (Comment on them is coming later.)


Solution


This has a bunch of observations at each dose (100 cells, in
fact), so we need to do the two-column response thing: the
successes in the first column and the failures in the
second. This means that we need to calculate the number of
cells at each dose that were not damaged, by subtracting the
number that *were* damaged from 100. This is best done within the `glm`, so that everything is accessible to `plot_predictions` later.



Now we fit our model:

```{r doseresponse-8 }
cells.1 <- glm(cbind(damaged, 100 - damaged) ~ dose, family = "binomial", data = dr)
summary(cells.1)
```

 
The programmer in you is probably complaining "but, 100 is a number and `damaged` is a vector of 7 numbers. How does R know to subtract *each one* from 100?" Well, R has what's known as
"recycling rules": if you try to add or subtract (or elementwise
multiply or divide) two vectors of different lengths, it recycles the
smaller one by repeating it until it's as long as the longer one. So
rather than `100-damaged` giving an error, it does what you
want.^[The usual application of this is to combine a number  with a vector. If you try to subtract a length-2 vector from a  length-6 vector, R will repeat the shorter one three times and do  the subtraction, but if you try to subtract a length-2 vector from a length-*7* vector, where you'd have to repeat the shorter one a fractional number of times, R will do it, but you'll get a warning, because this is probably *not* what you wanted. Try it and see.]

    

$\blacksquare$

(d) Does your output indicate that the probability of damage
really does *increase* with dose? (There are two things here:
is there really an effect, and which way does it go?)


Solution


The slope of `dose` is significantly nonzero (P-value
less than $2.2 \times 10^{-16}$, which is as small as it can
be). The slope itself is *positive*, which means that as
dose goes up, the probability of damage goes up. 

That's all I needed, but if you want to press on: the slope is
0.113, so an increase of 1 in dose goes with an increase of
0.113 in the *log-odds* of damage. Or it multiplies the
odds of damage by $\exp(0.113)$. Since 0.113 is small, this is
about $1.113$ (mathematically, $e^x\simeq 1+x$ if $x$ is small),
so that the increase is about 11\%. 
If you like, you can get a rough CI for the slope by going up
and down twice its standard error (this is the usual
approximately-normal thing). Here that would be

```{r doseresponse-9 }
0.113323 + c(-2, 2) * 0.008315
```

       

I thought about doing that in my head, but thought better of it, since
I have R just sitting here. The interval is short (we have lots of
data) and definitely does not contain zero.
    

$\blacksquare$

(e) Obtain predicted damage probabilities for some representative doses. 


Solution

Pick some representative doses first, and then use them as `newdata`. The ones in the original data are fine (there are only seven of them), but you can use some others if you prefer:

```{r}
cbind(predictions(cells.1, newdata = dr)) %>% 
  select(dose, damaged, estimate, conf.low, conf.high) -> p
p
```

I saved mine to use again later, but you don't have to unless *you* want to use your predictions again later.

$\blacksquare$

(f) Draw a graph of the predicted probabilities, and to that add the observed proportions of damage at each dose. Hints: you will have to calculate the observed proportions first. 
Looking at  the predicted probabilities, would you say that
the model fits well? Explain briefly.


Solution

Because we used the `cbind` thing in our `glm`, we can start with `plot_predictions` as expected:

```{r}
plot_predictions(cells.1, condition = "dose")
```

The predicted probability of a cell being damaged is very precisely estimated all the way up, and the probability is almost certainly very close to 1 for large doses.

Now, we need to figure out how to add the data to this graph.
Let's take a look at our original dataframe:

```{r}
dr
```

To that we need to add a column of proportion damaged, which is `damaged` divided by 100 (we know there were exactly 100 cells at each dose):

```{r}
dr %>% mutate(prop = damaged / 100) -> dr2
dr2
```

Check. I saved this to add to the graph.


Now you can add this to the graph we made before with a `geom_point` with a `data =` and an `aes`, making the points red:

```{r}
#| error: true

plot_predictions(cells.1, condition = "dose") +
  geom_point(data = dr2, aes(x = dose, y = prop), colour = "red")
```

I also made the data points red (you don't need to, but if you want to, put the colour *outside* the `aes`, to make it clear that the colour is constant, not determined by any of the variables in your dataframe).

The predicted
probabilities ought to be close to the observed proportions. They are in fact *very* close to them, so the model
fits very well indeed.

Your actual words are a judgement call, so precisely what you
say doesn't matter so much, but *I* think this model fit is
actually closer than you could even hope to expect, it's that
good. But, your call. I think your answer ought to contain
"close" or "fits well" at the very least. 
     
$\blacksquare$




