##  Predicting voting


 1257 British voters were classified according
to their social class, age (categorized), sex and the political party
they voted for (Labour or Conservative). Which, if any, of these
factors influences the party that someone votes for? The data are in
[link](http://ritsokiguess.site/datafiles/voting.txt), one voter
per line.



(a) Read in the data and display (some of) the data frame.

Solution


Space-delimited:

```{r vote-1 }
my_url <- "http://ritsokiguess.site/datafiles/voting.txt"
vote0 <- read_delim(my_url, " ")
vote0
```

     

I gave it a "disposable" name, since we make the "real" data set
shortly. 

$\blacksquare$

(b) There is no frequency column here, because each row of the
data frame only represents one voter. Count up the frequencies for
each combo of the categorical variables, and save it (this is the
data frame that we will use for the analysis).
Display the first few rows of the result. Do you now
have something that you need?

Solution


I changed my mind about how to do this from last year. Using
`count` is alarmingly more direct than the method I had before:

```{r vote-2 }
votes <- vote0 %>% count(class, age, sex, vote)
votes
```

     

Exactly the right thing now: note the new column `n` with
frequencies in it. (Without a column of frequencies we can't fit a
log-linear model.) There are now only 58 combinations of the four
categorical variables, as opposed to 1247 rows in the original data
set (with, inevitably, a lot of repeats).

$\blacksquare$

(c) Fit a log-linear model with the appropriate interaction (as a
starting point).

Solution


```{r vote-3 }
vote.1 <- glm(n ~ class * age * sex * vote, data = votes, family = "poisson")
```

     

$\blacksquare$

(d) Refine your model by taking out suitable non-significant
terms, in multiple steps. What model do you finish with?

Solution


Alternating `drop1` and `update` until everything
remaining is significant:

```{r vote-4 }
drop1(vote.1, test = "Chisq")
```

     

Not anywhere near significant, so out it comes:

```{r vote-5 }
vote.2 <- update(vote.1, . ~ . - class:age:sex:vote)
drop1(vote.2, test = "Chisq")
```

 

Take out the first one, since it has the highest P-value:

```{r vote-6 }
vote.3 <- update(vote.2, . ~ . - class:age:sex)
drop1(vote.3, test = "Chisq")
```

 

`class:sex:vote`:

```{r vote-7 }
vote.4 <- update(vote.3, . ~ . - class:sex:vote)
drop1(vote.4, test = "Chisq")
```

 

`class:sex`:

```{r vote-8 }
vote.5 <- update(vote.4, . ~ . - class:sex)
drop1(vote.5, test = "Chisq")
```

 

I don't like having three-way interactions, so I'm going to yank
`age:sex:vote` now, even though its P-value is smallish:

```{r vote-9 }
vote.6 <- update(vote.5, . ~ . - age:sex:vote)
drop1(vote.6, test = "Chisq")
```

 

The age-sex interaction can go, but we must be near the end now:

```{r vote-10 }
vote.7 <- update(vote.6, . ~ . - age:sex)
drop1(vote.7, test = "Chisq")
```

 

And that's it. The age and sex main effects are not included in the
list of droppable things because
both variables are part of higher-order interactions that are still in
the model.

If you want to, you can look at the `summary` of your final model:

```{r vote-11, echo=F}
options(width = 90)
```

 

```{r vote-12, size="footnotesize"}
summary(vote.7)
```

 

These tend to be rather unwieldy, and we'll see a better way of
understanding the results below, but you can look for the very
significant results, bearing in mind that the first category is the
baseline, for example, more of the males in the survey voted Labour
(than Conservative).

$\blacksquare$

(e) If we think of the party someone votes for as the final
outcome (that depends on all the other things), what does our final
model say that someone's vote depends on?

Solution


Find out which of the surviving terms are interactions with
`vote`. Here, there are two things, that `vote`
depends on separately:


* `sex`

* The `age`-`class` interaction.


$\blacksquare$

(f) Obtain graphs that help you explain how `vote` depends on
any of the things it's related to.

Solution

The 3-way interaction is a bit
tricky, so we'll do the simpler one first:

```{r}
ggplot(votes, aes(x = sex, y = n, fill = vote)) + 
  geom_col(position = "fill")
```


The female voters slightly preferred to vote Conservative^[The colours came out the wrong way around: the Conservative colour is blue and the Labour colour is red, like the Conservatives and Liberals here.] and the male
voters slightly preferred to vote Labour. This is a small effect, but
I guess the large number of voters made it big enough to be significant.

I took it this way around because `vote` is the outcome, and
therefore I want to address things 
like "if a voter is female, how likely are they to vote Labour", 
rather than conditioning the other
way around (which would 
be "if a voter voted Labour, how likely are they to be female", 
which doesn't make nearly so much sense). 

Then the tricky one. For this, we need to choose one  of the explanatory variables to be `x` and the other one to be `facets`. There is no particular reason to do it one way around rather than the other; experiment with switching `x` and the variable in `facet_wrap` around and see whether it's easier to interpret:

```{r}
ggplot(votes, aes(x = age, y = n, fill = vote)) +
  geom_col(position = "fill") + facet_wrap(~ class)
```

The age and class categories are in an illogical alphabetical order, but they were out of order in the original dataset too, so I guess we have to live with what we see here.

Doing it this way has produced different graphs for each
`class`. This is actually OK, because we can 
say "if a voter was of lower middle class" and then talk about the relationship
between age and vote, as if we were looking at a simple effect:



* If a voter was of lower-middle-class, they strongly favour voting
Conservative in all age groups except for `<26` and 26--40.

* If a voter was of upper-middle-class, they even more strongly favour
voting Conservative in all age groups except for "under 26" and
maybe 26--40.

* If a voter was of Working class, they strongly favour voting
Labour, except in the 
"over 75" age group (and maybe 51--75 as well).


If the anomalous age group(s) had been the same one every time, there
would no longer have been an interaction between age and class in
their effect on `vote`. But the anomalous age groups were
different for each class ("different pattern"), and that explains
why there was a `vote:age:class` interaction: " the way someone votes depends on the *combination* of age and social class". 

Extra 1: I have been using the dataframe with frequencies in it, and telling you to use `geom_col` rather than the `geom_bar` you might have been expecting, which is because we have frequencies and each row of the dataframe represents a bunch of people rather than just one. But let's remind ourselves of the dataframe we had called vote0:

```{r}
vote0
```

This has exactly one voter per row (there were 1257 people in the original survey), and so with this one you *can* use `geom_bar` to get the same graphs we got above using `geom_col`. For example:

```{r}
ggplot(vote0, aes(x = age, fill = vote)) +
  geom_bar(position = "fill") + facet_wrap(~ class)
```

`geom_bar` doesn't allow a `y`, but `vote0` doesn't have the frequency column that we were using for the `y`, so we just take it out.

Extra 2: The Labour Party in the UK is like the NDP here, in that it has strong
ties with "working people", trades unions in particular. The
Conservatives are like the Conservatives here (indeed, the nickname
"Tories" comes from the UK; the Conservatives there were officially
known as the Tories many years ago). Many people are lifelong voters
for their party, and would never think of voting for the "other side", 
in the same way that many Americans vote either Democrat or
Republican without thinking about it too much. Our parliamentary
system comes from the UK system (vote for a candidate in a riding, the
leader of the party with the most elected candidates becomes Prime
Minister), and a "landslide" victory often comes from persuading
enough of the voters open to persuasion to switch sides. In the UK, as
here, the parties' share of the popular vote doesn't change all that
much from election to election, even though the number of seats in
Parliament might change quite a lot. Provincial elections in Canada work the same way; you might remember that the current (as of 2022) Conservative majority in Ontario came from less than 40% of the vote.

$\blacksquare$

